### 2020-09-10 Manuscript/manuscript.tex
add Proterozoic evaporite text

Extensive rifting followed by thermal subsidence occurred along the southeast to east Laurentia margin in the time leading up to the Neoproterozoic-Phanerozoic boundary and is interpreted to be associated with the opening of the Iapetus ocean. A record of this rifting is preserved as rift basins that were part of failed arms (Rome trough, Reelfoot rift and Oklahoma aulacogen; Fig. \ref{fig:Laurentia_map}) as well as prolonged Cambrian to Ordovician passive margin thermal subsidence along the margin \citep{Bond1984a, Whitmeyer2007a}. The age of igneous intrusions that have been interpreted to be rift-related play a significant role in interpretations of this history such as in the rift development model of \citet{Burton2010a}. In this model, spatially-restricted rifting occurs ca. 760 to 680 Ma in the region of modern-day North Carolina and Virginia. Rifting ca. 620 to 580 Ma initiates in the region from modern-day New York to Newfoundland and by ca. 580 to 550 Ma rifting extends along the length of Laurentia's eastern margin. The last phase of this rifting has been interpreted to be associated with the separation of the Argentine pre-Cordillera Cuyania terrane \citep{Dickerson1998a}. As with other rifts, it is difficult to distinguish the separation of a cratonic fragment as a microcontinent from the rifting and departure of a major craton, as the record that lingers on the craton is similar. Recognizing this ambiguity, \citet{Robert2020a} propose that rather than being associated with spatially-restricted or failed rifting that ca. 700 Ma extension is associated with breakup and separation of Laurentia and its conjugate continent (that they interpret to be Amazonia). This rifting would have led to the formation of the Paleo-Iapetus Ocean (an analogue with the Paleo-Thethys). Subsequent to this rifting of the major continental blocks, smaller terranes of lithosphere rift off the east Laurentia ca. 600 Ma leading to the formation of the Neo-Iapetus Ocean and the record of passive margin development on Laurentia \citep{Robert2020a}.

@@

%A nice touch would be to add Paleoproterozoic, Meso, Neo and Phanero onto this figure
\begin{figure}
\centering
\includegraphics[width=\textwidth]{../Figures/Laurentia_paleolatitude.pdf}
\caption{\small{\textbf{Top panel: Paleolatitude implied by paleomagnetic poles from Laurentia and associated blocks for Duluth (lat=46.79\textdegree N, lon=92.10\textdegree W). The paleomagnetic poles are compiled in Table 2. Bottom panel: Paleolatitude implied by Laurentia poles compared with that implied by published paleogeographic models and the simple Laurentia model used in this chapter for the reconstructions in Figure \ref{fig:Laurentia_reconstructions}.}}}
\label{fig:Laurentia_paleolatitude}
\end{figure}

@@

@@

\subsection{Evaluating Laurentia's Proterozoic paleogeographic neighbors}
\textit{Note that this section is currently under development and is incomplete}

@@

In the central and southern Appalachians there are inliers of rocks that were metamorphosed during the Ottawan phase of the Grenvillian orogeny \citep{McLelland2013a}. On the basis of whole-rock Pb-isotope data, \cite{Loewy2003a} and \cite{Fisher2010a} proposed that these inliers are fragments of Amazonia lithosphere that were transferred to Laurentia during the orogeny and left behind when the Iapetus Ocean formed. In particular, \cite{Fisher2010a} suggest the Suns\'as orogen of Amazonia as the best match for southern and central Appalachian inliers. This positioning leads to a paleogeographic model wherein Amazonia is a major portion of the conjugate continental lithosphere that collided with Laurentia during Rodinia assembly \citep{Evans2013a,Cawood2017a}. In this model, the basement inliers of the Appalachian Orogen in the Blue Ridge region are interpreted to be the leading edge of Amazonia with initial collision ca. 1080 initiating the Ottawan phase of the Grenvillian orogeny (Fig. \ref{fig:tectonic_history}). Subsequent separation of Amazonia would have lead to the formation of the Iapetus Ocean as Rodinia rifted apart. Departure of Amazonia potentially occurred as early as ca. 700 Ma in the Paleo-Iapetus Ocean model of  \citet{Robert2020a}.

%%\subsubsection{Baltica}
%%
\subsubsection{North China}

>>>>

Extensive rifting followed by thermal subsidence occurred along the southeast to east Laurentia margin in the time leading up to the Neoproterozoic-Phanerozoic boundary and is interpreted to be associated with the opening of the Iapetus ocean. A record of this rifting is preserved as rift basins that were part of failed arms (Rome trough, Reelfoot rift and Oklahoma aulacogen; Fig. \ref{fig:Laurentia_map}) as well as prolonged Cambrian to Ordovician passive margin thermal subsidence along the margin \citep{Bond1984a, Whitmeyer2007a}. The age of igneous intrusions that have been interpreted to be rift-related play a significant role in interpretations of this history such as in the rift development model of \citet{Burton2010a}. In this model, spatially-restricted rifting occurs ca. 760 to 680 Ma in the region of modern-day North Carolina and Virginia. Rifting ca. 620 to 580 Ma initiates in the region from modern-day New York to Newfoundland and by ca. 580 to 550 Ma rifting extends along the length of Laurentia's eastern margin. The last phase of this rifting has been interpreted to be associated with the separation of the Argentine pre-Cordillera Cuyania terrane \citep{Dickerson1998a}. As with other rifts, it is difficult to distinguish the separation of a cratonic fragment as a microcontinent from the rifting and departure of a major craton, as the record that lingers on the craton is similar. Recognizing this ambiguity, \citet{Robert2020a} propose that rather than being associated with spatially-restricted or failed rifting that ca. 700 Ma extension is associated with breakup and separation of Laurentia and its conjugate continent (that they interpret to be Amazonia). This rifting would have led to the formation of the Paleo-Iapetus Ocean (an analogue with the Paleo-Tethys). Subsequent to this rifting of the major continental blocks, smaller terranes of lithosphere rift off the east Laurentia ca. 600 Ma leading to the formation of the Neo-Iapetus Ocean and the record of passive margin development on Laurentia \citep{Robert2020a}.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../Figures/Laurentia_paleolatitude.pdf}
\caption{\small{\textbf{Top panel: Paleolatitude implied by paleomagnetic poles from Laurentia and associated blocks for Duluth (lat=46.79\textdegree N, lon=92.10\textdegree W). The paleomagnetic poles are compiled in Table 2. Bottom panel: Paleolatitude implied by Laurentia poles compared with that implied by published paleogeographic models and the simple Laurentia model used in this chapter for the reconstructions in Figure \ref{fig:Laurentia_reconstructions}.}}}
\label{fig:Laurentia_paleolatitude}
\end{figure}

@@

\subsection{Paleoenvironmental constraints on paleolatitude}

Sedimentary rocks whose deposition is associated with certain climatic conditions have the potential to provide insight into paleolatitude. 
Relevant deposits in the Proterozoic include glacial deposits deposited by continental ice sheets, carbonates deposited in carbonate-saturated (and thereby likely to be warm) marine environments, and evaporite deposits deposited where evaporation exceeded precipitation. Interpretations of paleolatitude based on glacial deposits are complicated by the evidence for multiple global and low-latitude glacial intervals during the Proterozoic \citep{Evans2003a}. Evaporite deposits are particularly compelling as a paleolatitude constraint given that their deposition is associated arid regions resulting from large scale Hadley cell downwelling \citep{Evans2006a}. While moisture in the subtropics can change along with Earth's climate the overall pattern of $\sim$15-35\textdegree$\;$ of latitude being where annual mean evaporation exceeds precipitation persists \citep{Burls2017a}. Using compiled data that paired paleomagnetically-determined paleolatitude constraints and evaporite occurrence, \cite{Evans2006a} demonstrated that over the past 2 Ga that large-scale evaporite deposition was consistently located in subtropical latitudes that correspond to modern arid zones. This finding is consistent both with the geocentric axial dipole hypothesis used to calculate paleolatitude and the long-term stability of large-scale Hadley cell circulation.

Proterozoic evaporite deposits are documented within these units that were deposited following the amalgamation of Laurentia: 
\begin{itemize}
%\item The Stark Formation of the Slave Province contains displacive halite pseudomorphs throughout as well as massive breccias  interpreted to be evaporite-collapse breccias \citep{Pope2003a}. The formation is intruded by 1865 $\pm$ 15 Compton laccoliths  and is interpreted to have been deposited 
\item The Altyn Formation of the Belt Supergroup contains pseudomorphs after gypsum crystals and anhydrite within shallow-water carbonates with relict preserved gypsum and anhydrite within secondary silica \citep{White1984a}. The correlative Prichard Formation is 
intruded by 1468.8 $\pm$ 2.5 Ma sills \citep{Sears1998a}. Halite molds and casts are present within mudstones of the overlying Grinell Formation \citep{Pratt2019a}. Further up in the Belt Supergroup , within the Wallace Formation, there is stratiform scapolite --- a metamorphic mineral interpreted to have formed from a halite precursor within the Wallace Formation \citep{Hietanen1967a}. These deposits are older than the 1443 $\pm$ 7 Ma Purcell lavas and likely similar in age to the 1454 $\pm$ 9 Ma tuff within Helena Formation \citep{Evans2000c}. 
\item The Mesoproterozoic Iqqittuq Formation of the Borden basin (formerly part of the Society Cliffs Formation) contains bedded gypsum deposits (massive and laminated) and shale with halite casts \citep{Kah2001a}.  These deposits are bracketed between Re-Os dates of 1048 $\pm$ 12 Ma for an underlying shale and 1046 $\pm$ 16 Ma for an overlying shale \citep{Gibson2018a}  While gypsum beds can reach thicknesses of 2.5 meters \citep{Kah2001a}, evaporites in the Iqqittuq Formation are interpreted to be volumetrically minor compared to units such as the Tonian Ten Stone Formation \citep{Turner2009a}.
\item the Tonian Ten Stone Formation of the Mackenzie Mountains Supergroup (formerly known as the Gypsum Formation) contains a $\sim$500 meter thick successions dominated by gypsum with minor anhydrite interpreted to have been deposited in a deep-water (below wave base) restricted marine basin \citep{Turner2016a} . These thick bedded sulfate deposits are older than cross-cutting 777.7 $\pm$ 2.5 Ma sills of the Gunbarrel large igneous province (U-Pb date from \citealp{Jefferson1989a}) and younger than ca. 1005 Ma detrital zircons \citep{Turner2016a}. The overlying Ram Head Formation has been correlated with the Bitter Springs Stage which is constrained between 811.5 $\pm$ 0.3 Ma and 788.7 $\pm$ 0.2 Ma \citep{Macdonald2010a, Swanson-Hysell2015a} suggesting that the evaporites are ca. 820 Ma \citep{Turner2016a}. These deposits are hypothesized to be correlative with sulfate evaporites within the Minto Inlet Formation of the Shaler Supergroup \citep{Jones2010a, Turner2016a}.
\item the Tonian Kilian Formation of the Shaler Supergroup contains nodules of gypsum and anhydrite interpreted to have been deposited in intertidal to supratidal evaporitic mudflat environment \citep{Prince2014a}. The Kilian Formation is interpreted to post-date the Bitter Springs Stage and be correlative with the Redstone River Formation of the Coates Lake Group in the McKenzie Mountains that contains bedded gypsum as well as gypsum bearing siltstone\citep{Jefferson1989a, Jones2010a}. The Redstone River Formation is younger than the 777.7 $\pm$ 2.5 Ma volcanics and older than a 732.2 $\pm$ 4.7 Ma Re-Os isochron from the overlying Coppercap Formation.
\end{itemize}

@@

\subsection{Evaluating Laurentia's Proterozoic paleogeographic neighbors}
Many different paleogeographic connections between Laurentia and other Proterozoic cratons that have been proposed and utilized in paleogeographic models. This section does not seek to be comprehensive in this regard, but rather I seek to highlight and contextualize some of the more prominent and/or well-supported models.
\textit{Note that this section is currently under development and is incomplete}

@@

In the central and southern Appalachians there are inliers of rocks that were metamorphosed during the Ottawan phase of the Grenvillian orogeny \citep{McLelland2013a}. On the basis of whole-rock Pb-isotope data, \cite{Loewy2003a} and \cite{Fisher2010a} proposed that these inliers are fragments of Amazonia lithosphere that were transferred to Laurentia during the orogeny and left behind when the Iapetus Ocean formed. In particular, \cite{Fisher2010a} suggest the Suns\'as orogen of Amazonia as the best match for southern and central Appalachian inliers. This positioning leads to a paleogeographic model wherein Amazonia is a major portion of the conjugate continental lithosphere that collided with Laurentia during Rodinia assembly \citep{Evans2013a, Cawood2017a}. While the lack of ca. 1100 and ca. 1000 Ma poles from Amazonia precludes a robust paleomagnetic test, this scenario is consistent with the available late Mesoproterozoic poles from Amazonia (ca. 1200 Nova Floresta pole and ca. 1150 Fortuna Formation pole; D'Agrella-Filho et al., this volume) as shown in \cite{Evans2013a}. In this paleogeographic scenario, the basement inliers of the Appalachian Orogen in the Blue Ridge region are interpreted to be the leading edge of Amazonia with initial collision ca. 1080 initiating the Ottawan phase of the Grenvillian orogeny (Fig. \ref{fig:tectonic_history}). Subsequent separation of Amazonia would have lead to the formation of the Iapetus Ocean as Rodinia rifted apart. Departure of Amazonia potentially occurred as early as ca. 700 Ma in the Paleo-Iapetus Ocean model of  \citet{Robert2020a}.

\subsubsection{Baltica}
Based on correlation of Archean provinces and Paleoproterozoic orogenic belts, \cite{Gower1990a} reconstructed Baltica to Laurentia in a position known as the NENA (northern Europe and North America) configuration. This connection proposes a tight fit between modern-day northern Norway and Russia's Kola Peninsula with eastern Greenland (\citealp{Gower1990a}, Salminen et al., this volume). In this position, Baltica and Laurentia are hypothesized to share a long-lived accretionary margin \citep{Karlstrom2001a}. This reconstruction is brings paleomagnetic poles

compatible with existing paleomagnetic constraints from ca. 1750 to 1270 Ma \cite{Evans2008a} and these conjoined cratons feature as a major component of the hypothesized Nuna supercontinent (\citealp{Evans2011a, Zhang2012a}; .

\subsubsection{North China}
